**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-11-19 paper/manuscript.Rmd
add goldilocks citation

### Abstract
Colorectal cancer is a common and deadly disease in the United States accounting for over 50,000 deaths in 2020. This progressive disease is highly preventable with early detection and treatment, but many people do not comply with the recommended screening guidelines. The gut microbiome has emerged as a promising target for non-invasive detection of colorectal cancer. Most microbiome-based classification efforts utilize taxonomic abundance data from operational taxonomic units (OTUs) or amplicon sequence variants (ASVs) with the goal of increasing taxonomic resolution. However, it is unknown which taxonomic resolution is optimal for microbiome-based classification of colorectal cancer. To address this question, we used a reproducible machine learning framework to quantify classification performance of models based on data annotated to phylum, class, order, family, genus, OTU, and ASV levels. We found that model performance increased with increasing taxonomic resolution, up to the family level where performance was equal (p > 0.05) among family (mean AUC: `r format_decimal(rf_mean_auc$family,digits=3)`), genus (mean AUC: `r format_decimal(rf_mean_auc$genus,digits=3)`), and OTU (mean AUC: `r format_decimal(rf_mean_auc$otu,digits=3)`) levels before decreasing at the ASV level (p < 0.05, mean AUC: `r format_decimal(rf_mean_auc$asv,digits=3)`). These results demonstrate a trade-off between taxonomic resolution and prediction performance, where coarse taxonomic resolution (e.g. phylum) is not distinct enough but fine resolution (e.g. ASV) is to individualized to accurately classify samples. Similar to the story of Goldilocks and the three bears, mid-range resolution (i.e. family, genus, and OTU) is just right for optimal prediction of colorectal cancer from microbiome data.

### Importance
Despite being highly preventable, colorectal cancer remains a leading cause of cancer related death in the United States. Low-cost, non-invasive detection methods could greatly improve our ability to identify and treat early stages of disease. The microbiome has shown promise as a resource for detection of colorectal cancer. Research on the gut microbiome tends to focus on improving our ability to profile species and strain level taxonomic resolution. However, we found that finer resolution impedes the ability to predict colorectal cancer based on the gut microbiome. These results highlight the need for consideration of the appropriate taxonomic resolution for microbiome analyses and that finer resolution is not always more informative.

@@

While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUROC can have different ROC curve shapes [@lobo2008]. Depending on the intended implementation of the model, one may want to optimize the sensitivity over the specificity or vice versa. In this case, the optimal model will detect as many true positives (people with SRNs) as possible. To further compare the model performance across taxonomic levels we compared the sensitivity of the models at a specificity of 90%. The highest sensitivity values were observed for family (mean sensitivity: `r format_decimal(mean_sens_spec90$family,digits=2)`), genus (mean sensitivity: `r format_decimal(mean_sens_spec90$genus,digits=2)`), and OTU (mean sensitivity: `r format_decimal(mean_sens_spec90$otu,digits=2)`) level data (p > 0.05, Figure 1B), consistent with the AUROC results. Phylum (mean sensitivity: `r format_decimal(mean_sens_spec90$phylum,digits=2)`), class (mean sensitivity: `r format_decimal(mean_sens_spec90$class,digits=2)`), order (mean sensitivity: `r format_decimal(mean_sens_spec90$order,digits=2)`), and ASV (mean sensitivity: `r format_decimal(mean_sens_spec90$asv,digits=2)`) sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (p < 0.05,Figure 1B). This analysis further supports the observation that finer resolution does not improve SRN detection.

@@

***Data processing.*** Sequence data was processed with mothur (1.44.3) [@schloss2009] using the SILVA reference database (v132) [@quast2013] to produce count tables for phylum, class, order, family, genus, OTU, and ASV following the Schloss Lab MiSeq SOP described on the mothur website (<https://mothur.org/wiki/miseq_sop/>). ASV level data was also produced using DADA2 [@callahan2016] to ensure consistent results with a different pipeline. Data was processed following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from the whole dataset rather than per sample. The resulting ASV table was subsampled for consistency with the mothur data. The DADA2 generated ASVs performed worse than the mothur generated ASVs (DADA2 ASV mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`, p < 0.05).

@@

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent

<div id="refs"></div>

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

>>>>

### Abstract

Colorectal cancer is a common and deadly disease in the United States accounting for over 50,000 deaths in 2020. This progressive disease is highly preventable with early detection and treatment, but many people do not comply with the recommended screening guidelines. The gut microbiome has emerged as a promising target for non-invasive detection of colorectal cancer. Most microbiome-based classification efforts utilize taxonomic abundance data from operational taxonomic units (OTUs) or amplicon sequence variants (ASVs) with the goal of increasing taxonomic resolution. However, it is unknown which taxonomic resolution is optimal for microbiome-based classification of colorectal cancer. To address this question, we used a reproducible machine learning framework to quantify classification performance of models based on data annotated to phylum, class, order, family, genus, OTU, and ASV levels. We found that model performance increased with increasing taxonomic resolution, up to the family level where performance was equal (p \> 0.05) among family (mean AUC: `r format_decimal(rf_mean_auc$family,digits=3)`), genus (mean AUC: `r format_decimal(rf_mean_auc$genus,digits=3)`), and OTU (mean AUC: `r format_decimal(rf_mean_auc$otu,digits=3)`) levels before decreasing at the ASV level (p \< 0.05, mean AUC: `r format_decimal(rf_mean_auc$asv,digits=3)`). These results demonstrate a trade-off between taxonomic resolution and prediction performance, where coarse taxonomic resolution (e.g. phylum) is not distinct enough but fine resolution (e.g. ASV) is to individualized to accurately classify samples. Similar to the story of Goldilocks and the three bears[@cauley1981], mid-range resolution (i.e. family, genus, and OTU) is just right for optimal prediction of colorectal cancer from microbiome data.

### Importance

Despite being highly preventable, colorectal cancer remains a leading cause of cancer related death in the United States. Low-cost, non-invasive detection methods could greatly improve our ability to identify and treat early stages of disease. The microbiome has shown promise as a resource for detection of colorectal cancer. Research on the gut microbiome tends to focus on improving our ability to profile species and strain level taxonomic resolution. However, we found that finer resolution impedes the ability to predict colorectal cancer based on the gut microbiome. These results highlight the need for consideration of the appropriate taxonomic resolution for microbiome analyses and that finer resolution is not always more informative.

@@

While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUROC can have different ROC curve shapes [@lobo2008]. Depending on the intended implementation of the model, one may want to optimize the sensitivity over the specificity or vice versa. In this case, the optimal model will detect as many true positives (people with SRNs) as possible. To further compare the model performance across taxonomic levels we compared the sensitivity of the models at a specificity of 90%. The highest sensitivity values were observed for family (mean sensitivity: `r format_decimal(mean_sens_spec90$family,digits=2)`), genus (mean sensitivity: `r format_decimal(mean_sens_spec90$genus,digits=2)`), and OTU (mean sensitivity: `r format_decimal(mean_sens_spec90$otu,digits=2)`) level data (p \> 0.05, Figure 1B), consistent with the AUROC results. Phylum (mean sensitivity: `r format_decimal(mean_sens_spec90$phylum,digits=2)`), class (mean sensitivity: `r format_decimal(mean_sens_spec90$class,digits=2)`), order (mean sensitivity: `r format_decimal(mean_sens_spec90$order,digits=2)`), and ASV (mean sensitivity: `r format_decimal(mean_sens_spec90$asv,digits=2)`) sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (p \< 0.05,Figure 1B). This analysis further supports the observation that finer resolution does not improve SRN detection.

@@

***Data processing.*** Sequence data was processed with mothur (1.44.3) [@schloss2009] using the SILVA reference database (v132) [@quast2013] to produce count tables for phylum, class, order, family, genus, OTU, and ASV following the Schloss Lab MiSeq SOP described on the mothur website (<https://mothur.org/wiki/miseq_sop/>). ASV level data was also produced using DADA2 [@callahan2016] to ensure consistent results with a different pipeline. Data was processed following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from the whole dataset rather than per sample. The resulting ASV table was subsampled for consistency with the mothur data. The DADA2 generated ASVs performed worse than the mothur generated ASVs (DADA2 ASV mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`, p \< 0.05).

@@

```{=tex}
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
```
\noindent

::: {#refs}
:::

```{=tex}
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
```
\newpage
